%% ============================================================
\section{Paillier Encryption}
%% ============================================================

\todo{To include a leading example of homomorphism of RSA and but it only brings about
multiplicative, then talks about GM encryption and its homomorphism, and finally some
intuition why paillier construction brings additive}
\todo{I think it may be in the GM paper, to check how to reduce from semantic security to IND-CPA and vice versa. }

Paillier encryption~\cite{Paillier1999} is an additively homomorphic public-key scheme based on the
\emph{decisional composite residuosity} (DCR) assumption.

%% ------------------------------------------------------------
\subsection{Setup and Key Generation}

The key generation algorithm proceeds as follows.

\begin{enumerate}
  \item Sample large primes $p, q$ satisfying $\gcd(pq,\, (p-1)(q-1)) = 1$, and set $n = pq$.
  \item Compute $\lambda \gets \mathrm{lcm}(p-1,\, q-1)$.
  \item Sample $g \gets \Z_{n^2}^*$ uniformly at random.
  \item Compute $\mu \gets \bigl(L(g^\lambda \bmod n^2)\bigr)^{-1} \bmod n$, where
        $L(x) \coloneqq \lfloor (x-1)/n \rfloor$.
        (This requires $L(g^\lambda \bmod n^2)$ to be invertible mod $n$; resample $g$ if not.)
  \item Return $pk \gets (n, g)$ and $sk \gets (\lambda, \mu)$.
\end{enumerate}

\begin{remark}[Simplified generator]
  The common choice $g = 1 + n$ is always valid and simplifies both key generation and
  decryption. Since $(1+n)^m \equiv 1 + mn \pmod{n^2}$ (Lemma~\ref{lem:binomial} below),
  one obtains $L(g^\lambda \bmod n^2) = \lambda \bmod n$, so $\mu = \lambda^{-1} \bmod n$
  is computable directly without sampling. The security level is identical; see
  Remark~\ref{rem:g-security}.
\end{remark}

\paragraph{Encryption.}
$\mathsf{Enc}(pk, m \in \Z_n)$: sample $r \gets \Z_n^*$ and return
\[
  c \;\equiv_{n^2}\; g^m \cdot r^n.
\]

\paragraph{Decryption.}
$\mathsf{Dec}(sk, c)$: compute
\[
  m \;\gets\; L\!\left(c^\lambda \bmod n^2\right) \cdot \mu \bmod n.
\]

\paragraph{Correctness.}
We use the following key lemma.

\begin{lemma}[Binomial structure of $\langle 1+n \rangle$]
\label{lem:binomial}
  For any $w \in \Z_{n^2}^*$, write its unique factorization in the direct product
  $\Z_{n^2}^* \cong \langle 1+n \rangle \times H$ (where $H = \{r^n : r \in \Z_n^*\}$ is
  the subgroup of $n$-th powers) as $w = (1+n)^{[w]} \cdot h$ for $[w] \in \Z_n$ and $h \in H$.
  Then:
  \begin{enumerate}
    \item $(1+n)^k \equiv 1 + kn \pmod{n^2}$ for all $k \in \Z$.
    \item $h^{n\lambda} \equiv 1 \pmod{n^2}$ for all $h \in H$.
    \item $L(w^\lambda \bmod n^2) = \lambda [w] \bmod n$.
  \end{enumerate}
\end{lemma}

\begin{proof}
  \textbf{(i)} By the binomial theorem,
  $(1+n)^k = \sum_{j=0}^{k} \binom{k}{j} n^j \equiv 1 + kn \pmod{n^2}$,
  since all terms with $j \ge 2$ are divisible by $n^2$.

  \textbf{(ii)} Any $h \in H$ has the form $h = r^n$ for $r \in \Z_n^*$. The element
  $r^n \in \Z_{n^2}^*$ has order dividing $\phi(n)$ (by Carmichael's theorem applied to
  $n^2$ together with the condition $\gcd(n, \phi(n)) = 1$), and $\lambda \mid \phi(n)$,
  so $h^\lambda = r^{n\lambda} \equiv 1 \pmod{n^2}$.

  \textbf{(iii)} Write $w = (1+n)^{[w]} \cdot h$. Then
  $w^\lambda = (1+n)^{[w]\lambda} \cdot h^\lambda = (1+n)^{[w]\lambda}$
  by (ii). Applying (i): $(1+n)^{[w]\lambda} \equiv 1 + [w]\lambda n \pmod{n^2}$,
  so $L(w^\lambda) = [w]\lambda \bmod n$.
\end{proof}

Correctness of decryption now follows immediately: $c = g^m r^n$ gives
$[c] = m[g]$ (since $[r^n] = 0$), so
\[
  L(c^\lambda) \cdot \mu
  = \lambda[c] \cdot \mu
  = \lambda m [g] \cdot (L(g^\lambda))^{-1}
  = \lambda m [g] \cdot (\lambda [g])^{-1}
  = m \bmod n.
\]

%% ------------------------------------------------------------
\subsection{The DCR Assumption}

\begin{tcolorbox}[colback=blue!4!white,colframe=blue!45!black]
  \textbf{DCR assumption.} Given $n = pq$ and a random $z \gets \Z_{n^2}^*$, no
  probabilistic polynomial-time algorithm can decide whether $z$ is an $n$-th power
  modulo $n^2$ with advantage non-negligibly greater than $\tfrac{1}{2}$.
\end{tcolorbox}

The $n$-th powers in $\Z_{n^2}^*$ form the subgroup $H = \{r^n \bmod n^2 : r \in \Z_n^*\}$
of index $n$ and order $\phi(n)$; the DCR assumption says the $n$ cosets of $H$ are
computationally indistinguishable.

\todo{State the assumption formally and sketch the reduction from DCR to IND-CPA security.}

\begin{remark}[Security with $g = 1+n$]
\label{rem:g-security}
  Using $g = 1+n$ does not weaken security. The ciphertext $c = (1+mn) \cdot r^n$ is
  uniformly random within the coset $(1+mn) \cdot H$. Under DCR, these cosets are
  computationally indistinguishable, so an adversary cannot determine $m$ from $c$
  regardless of which valid $g$ is used. The coset structure is determined entirely by
  $r^n$, not by the choice of $g$.
\end{remark}

%% ------------------------------------------------------------
\subsection{Additive Homomorphism}

For ciphertexts $c_1 = \mathsf{Enc}(m_1, r_1)$ and $c_2 = \mathsf{Enc}(m_2, r_2)$:
\[
  c_1 \cdot c_2 \bmod n^2
  = g^{m_1} r_1^n \cdot g^{m_2} r_2^n
  = g^{m_1+m_2} (r_1 r_2)^n
  = \mathsf{Enc}(m_1 + m_2 \bmod n,\; r_1 r_2).
\]
Scalar multiplication follows similarly:
\[
  c_1^k \bmod n^2 = g^{km_1}(r_1^k)^n = \mathsf{Enc}(k \cdot m_1 \bmod n,\; r_1^k).
\]
Both operations require only the public key and leave the randomness well-distributed.

%% ============================================================
\section{Threshold Paillier}
%% ============================================================

Threshold Paillier distributes the secret key $\lambda$ among $\ell$ parties so that any
$t$-of-$\ell$ subset can jointly decrypt, but no coalition of fewer than $t$ parties learns
anything about the plaintext.

%% ------------------------------------------------------------
\subsection{Motivation}

\todo{Explain why threshold decryption matters (MPC, auctions, e-voting, collaborative SNARKs).}

%% ------------------------------------------------------------
\subsection{Key Generation and Share Distribution}

We describe the \emph{trusted dealer} model; see Remark~\ref{rem:dkg} for the DKG variant.

\begin{enumerate}
  \item Run Paillier key generation to obtain $pk = (n, g)$ and $sk = (\lambda, \mu)$.
    Publish $pk$.
  \item Sample a random degree-$(t-1)$ polynomial $f \in \Z_{\phi(n)n}[X]$ with
    $f(0) = \phi(n)$ (so $sk = \phi(n)$).
    \todo{why this sharing is not in field? is it still perfectly secure?}
  \item For each party $i \in [\ell]$, set $sk_i \gets f(i) \bmod \phi(n)n$ and send $sk_i$
    privately to party $i$.
  \item Erase $f$ and $sk$.
\end{enumerate}

Set $\Delta \coloneqq \ell!$. For any subset $S \subseteq [\ell]$ with $|S| = t$ and any
$i \in S$, define the \emph{normalized Lagrange coefficient}
\[
  \Lambda_i(0) \;\coloneqq\; \Delta \cdot \lambda_i^S(0)
  \;=\; \Delta \prod_{j \in S \setminus \{i\}} \frac{-j}{i-j} \;\in\; \Z,
\]
which is always an integer since all denominators divide $\Delta = \ell!$.

\begin{remark}[DKG variant]
\label{rem:dkg}
To be completed
  % In the \emph{distributed key generation} (DKG) setting, no trusted dealer is assumed.
  % The parties jointly generate $(pk, sk_1, \dots, sk_\ell)$ via an MPC protocol such that
  % $sk$ never appears in the clear. For Paillier this is significantly harder than for
  % DL-based schemes, as it requires distributed generation of an RSA modulus $n = pq$
  % without any party knowing $p$ or $q$; see~\cite{BonehFranklin1997} and subsequent work.
  % In most practical deployments of threshold Paillier, a trusted dealer is assumed for
  % the setup phase.
\end{remark}

%% ------------------------------------------------------------
\subsection{Partial Decryption and Combining}

Given a ciphertext $c$ and a set $S$ of $t$ participating parties, decryption proceeds as
follows.

\paragraph{Partial decryption.}
Each party $i \in S$ computes its \emph{decryption share}:
\[
  c_i \;\equiv_{n^2}\; c^{2\Delta \cdot sk_i}.
\]
(The factor of $2$ is a technicality to ensure certain divisibility conditions; see
Shoup~\cite{Shoup2000}.)

\paragraph{Combining.}
The combiner computes:
\[
  \hat{m} \;\equiv_{n^2}\; \prod_{i \in S} c_i^{2\Lambda_i(0)}.
\]
Since $\sum_{i \in S} sk_i \cdot \Lambda_i(0) = \Delta \cdot sk$ by Lagrange interpolation,
the product evaluates to $c^{4\Delta^2 \cdot sk}$. Applying $L$ and scaling by the
appropriate precomputed constant recovers $m$.

\paragraph{Why integer exponents suffice.}
The Lagrange coefficients $\lambda_i^S(0)$ are rational in general. Multiplying by
$\Delta = \ell!$ clears all denominators (since differences $i - j$ for $i,j \in [\ell]$
divide $\ell!$), making $\Lambda_i(0) \in \Z$. This avoids the need to compute modular
inverses in $\Z_{n^2}^*$, which would require knowing $\phi(n^2)$ and hence the
factorization of $n$ --- precisely the secret we are trying to protect.

%% ------------------------------------------------------------
\subsection{Semi-Honest Security and Simulation}

The scheme above is secure against \emph{semi-honest} adversaries who corrupt up to $t-1$
parties and follow the protocol honestly but try to learn the plaintext from their views.
Security is proved by a reduction to plain Paillier IND-CPA security.

\paragraph{The reduction.}
The reduction is given a plain Paillier public key $N$ and access to a CPA challenge oracle.
It must simulate the entire threshold decryption game for an adversary $A$ who corrupts a
set $C \subseteq [\ell]$ with $|C| \leq t-1$.

\begin{enumerate}
  \item \textbf{Setup.} Sample $\Theta \leftarrow \Z_N$ and set $pk = (N, \Theta)$.
    For each corrupted party $i \in C$, sample $sk_i \leftarrow \Z_{N^2}$ uniformly and
    hand these to $A$. The reduction does not know the real $sk = \phi(N)$.

  \item \textbf{Decryption queries.} When $A$ asks for a decryption of message $m$, the
    reduction:
    \begin{enumerate}[(a)]
      \item Samples $r \leftarrow \Z_N$ and computes $c \equiv_{N^2} (1+mN) \cdot r^N$
        (possible using only $pk = N$).
      \item For each honest party $j \notin C$, computes the simulated share:
        \[
          c_j \;\equiv_{N^2}\;
          (1 + Nm\Theta)^{\Lambda_0(j)} \cdot \prod_{i \in C} c^{sk_i \cdot \Lambda_i(j)},
        \]
        where $\Lambda_0(j) = \Delta \prod_{k \in C} \frac{j-k}{0-k}$ and
        $\Lambda_i(j) = \Delta \cdot \frac{j}{i} \prod_{k \in C \setminus \{i\}} \frac{j-k}{i-k}$
        are normalized Lagrange coefficients for evaluating at $j$ from the set $C \cup \{0\}$.
    \end{enumerate}
    The key insight is that by Lemma~\ref{lem:binomial},
    $c^{sk} = (1+mN)^{\phi(N)} \equiv 1 + mN\phi(N) \equiv 1 + mN\Theta \pmod{N^2}$
    (since $\Theta = \phi(N) \bmod N$ in the real scheme, and here $\Theta$ plays this role).
    So the simulator expresses the honest party's contribution via $m$ and $\Theta$ alone,
    without needing $sk$.

  \item \textbf{Challenge.} $A$ sends $(m_0, m_1) \in \Z_N^2$. The reduction forwards
    $(\Theta\Delta^2 m_0 \bmod N,\ \Theta\Delta^2 m_1 \bmod N)$ to the plain Paillier CPA
    oracle and receives $c_b^*$. It forwards $c_b^*$ to $A$ as the challenge ciphertext,
    along with simulated shares computed as above. The messages are scaled by $\Theta\Delta^2$
    because the simulation introduces these extra factors: the combiner ultimately recovers
    $m\Delta^2\Theta \bmod N$ rather than $m$, so the challenge must be embedded accordingly.

  \item \textbf{Output.} $A$ outputs a guess $b'$; the reduction forwards it to the CPA
    challenger. Any advantage of $A$ translates directly to an advantage against plain
    Paillier IND-CPA.
\end{enumerate}

\begin{remark}[Why the simulation works]
  The simulated shares $c_j$ are indistinguishable from real ones because the honest
  party's share is determined by Lagrange interpolation --- from $A$'s view (which includes
  only the corrupted shares and the output), there is exactly one consistent value for each
  honest share, and the simulator produces exactly that value. No information about $sk$
  leaks beyond what is already implied by the corrupted shares and $m$.
\end{remark}

%% ------------------------------------------------------------
\subsection{Achieving Malicious Security via Verifiable Partial Decryptions}

The semi-honest scheme above fails against \emph{malicious} adversaries: a corrupt party
can submit an arbitrary $c_i' \neq c^{2\Delta \cdot sk_i}$ as its decryption share,
causing the combiner to reconstruct a bogus plaintext with no way to detect the fault.
To achieve malicious security, each partial decryption must come with a proof of
correctness.

\paragraph{Intuition.}
At setup the dealer publishes, alongside each share $sk_i$, a \emph{verification key}
$pk_i = v^{\Delta \cdot sk_i}$, where $v$ is a public generator of the subgroup
$\mathbb{Q}_{N^2}$ of quadratic residues modulo $N^2$.  When party $i$ submits
$c_i \equiv_{N^2} c^{2\Delta \cdot sk_i}$, any combiner can check consistency by
verifying that $c_i^2$ and $pk_i$ share the same discrete logarithm --- namely
$\Delta \cdot sk_i$ --- with respect to the bases $c^{4\Delta}$ and $v^{\Delta}$
respectively.  This is a \emph{proof of discrete-log equality (DLEQ)}: a
zero-knowledge argument that
\[
  \log_{c^{4\Delta}} c_i^2 \;=\; \log_{v^{\Delta}} pk_i \;=\; sk_i.
\]
A forged share $c_i' \neq c^{2\Delta \cdot sk_i}$ would require producing a valid DLEQ
proof for an incorrect relation, which is computationally infeasible under the
discrete-logarithm assumption in $\mathbb{Q}_{N^2}$.

\paragraph{Why safe primes and a special $g$?}
Standard DLEQ / Chaum--Pedersen proofs require working in a group of \emph{known} prime
order.  In $\mathbb{Z}_{N^2}^*$ the group order $\phi(N^2)$ is unknown (knowing it
factors $N$).  Two modifications address this:
\begin{enumerate}
  \item \textbf{Safe primes.} Choose $p = 2p'+1$ and $q = 2q'+1$ with $p', q'$ prime.
    Then $\mathbb{Q}_{N^2}$ has order $m = p'q' = N\phi(N)/4$, which is a product of two
    large primes.  Even though $m$ is unknown to the prover and verifier, the group has
    no small subgroups, which is the key property needed for the soundness argument.
  \item \textbf{Special generator.} Pick $g \equiv_{N^2} (1+N)^a b^N$ for $b \in \Z_N^*$
    and set $v = g^{2\alpha}$ for random $\alpha \in \Z_N^*$.  This ensures $v$ generates
    $\mathbb{Q}_{N^2}$ and, crucially, the reduction knows $\log_g v = 2\alpha$, which is
    needed to embed the DLEQ proof into a security argument against a plain-Paillier
    CPA adversary.
\end{enumerate}

\paragraph{Extra factors of $2$ and $4$.}
The exponent $2\Delta$ (rather than $\Delta$) in $c_i = c^{2\Delta \cdot sk_i}$ ensures
$c_i \in \mathbb{Q}_{N^2}$ (squaring maps any element into the QR subgroup), so that
DLEQ can be stated entirely within $\mathbb{Q}_{N^2}$.  The DLEQ proof itself introduces
a further squaring, so the combiner ultimately recovers $c^{4\Delta^2 \cdot sk}$ and must
divide out the extra factor of $4\Delta^2$ to obtain $m$.

\todo{
  Add the full description of the DLEQ proof used here. This is a zero-knowledge proof
  in the style of Chaum--Pedersen~\cite{ChaumPedersen1992}, adapted to a group of
  \emph{unknown} order. The key difference from the standard construction is that the
  challenge must be drawn from a sufficiently small set (e.g.\ $\{0,1\}^k$ for security
  parameter $k$) so that soundness holds even without knowing the group order; the
  completeness and zero-knowledge properties carry over unchanged. See
  Shoup~\cite{Shoup2000} for the full details and the soundness proof in this setting.
}

%% ------------------------------------------------------------
\subsection{Security}

\todo{
  State the threshold security game: semantic security holds as long as the adversary
  corrupts fewer than $t$ parties. Reference Shoup's 2000 practical threshold
  RSA/Paillier result.
}