\section{Groth16 and Pairing-Based SNARKs}
\label{sec:groth16}

The previous chapters developed the Plonk route to SNARKs: computation is represented as a trace, local constraints become polynomial identities, and wiring is enforced by a permutation argument. Groth16 follows a different but equally central route. It starts from a rank-1 constraint system (R1CS), converts that system into a quadratic arithmetic program (QAP), and then uses pairings to prove the resulting QAP relation succinctly \cite{GennaroGentryParnoRaykova2013,ParnoHowellGentryRaykova2013,Groth2016}.

This chapter stops at the arithmetization boundary:
\[
\text{algebraic circuit}\quad\longrightarrow\quad\text{R1CS}\quad\longrightarrow\quad\text{QAP}.
\]
The later pairing-based proving and verification equations of Groth16 are intentionally postponed.

\subsection{The arithmetization pipeline}

A high-level program is not directly suitable for a succinct proof. It must first be converted into algebraic statements over a field. For the Groth16 route, the pipeline is:
\[
\boxed{
\text{arithmetic circuit}
\longrightarrow
\text{flattened gates}
\longrightarrow
\text{R1CS matrices}
\longrightarrow
\text{QAP divisibility}
}
\]
The conceptual transformation is:
\[
\begin{array}{c}
\text{the circuit is evaluated correctly}\\[0.2em]
\Longleftrightarrow\quad \text{all rank-1 constraints are satisfied}\\[0.2em]
\Longleftrightarrow\quad P_z(r)=0\text{ for every constraint point }r\in H\\[0.2em]
\Longleftrightarrow\quad Z_H(X)\mid P_z(X).
\end{array}
\]
The last equivalence is the same algebraic principle that appeared in ZeroTest: if a polynomial is zero on a domain $H$, then it is divisible by the vanishing polynomial
\[
    Z_H(X)=\prod_{r\in H}(X-r).
\]

\subsection{From algebraic circuits to R1CS}

An arithmetic circuit over a field $\F$ consists of wires carrying field elements and gates computing additions or multiplications. A satisfying assignment contains public inputs, private witness values, and auxiliary wire values introduced during computation. We write these values as a single vector
\[
    z=(z_0,z_1,\ldots,z_{n-1})\in\F^n,
\]
where $z_0=1$ is reserved for constants. Appendix~\ref{app:r1cs-qap-example} gives a color-coded example in which variable colors track columns and constraint colors track rows, so that each R1CS matrix entry can be followed into its corresponding QAP term.

The first step is \emph{flattening}: every complex expression is decomposed into elementary equations whose left and right sides are linear forms, except for one multiplication between two linear forms. A general R1CS constraint has the shape
\begin{tcolorbox}
\[
    (\text{linear form in }z)\cdot(\text{linear form in }z)
    =
    (\text{linear form in }z).
\]
\end{tcolorbox}
Thus an addition gate such as
\[
    v=x_1+x_2
\]
can be written as
\[
    (x_1+x_2)\cdot 1=v,
\]
while a multiplication gate such as
\[
    y=v_1v_2
\]
is already of the form
\[
    v_1\cdot v_2=y.
\]

Formally, an R1CS instance with $m$ constraints and $n$ variables consists of three matrices
\[
    A,B,C\in\F^{m\times n}.
\]
The $i$th row of each matrix defines a linear form:
\[
    A_i\cdot z,
    \qquad
    B_i\cdot z,
    \qquad
    C_i\cdot z.
\]
The $i$th constraint is
\[
    (A_i\cdot z)(B_i\cdot z)=C_i\cdot z.
\]
Equivalently, all constraints can be written compactly as
\[
    (Az)\circ(Bz)=Cz,
\]
where $\circ$ denotes component-wise multiplication, not matrix multiplication. Each row is therefore one rank-1 multiplicative equation: the product of two linear forms equals a third linear form.

This representation is powerful because it separates the \emph{shape of the computation}, stored in $A,B,C$, from the \emph{specific execution}, stored in $z$. The public statement usually fixes some coordinates of $z$, while the prover supplies the private witness and auxiliary wire values.

\subsection{From R1CS to QAP}

R1CS is still a list of $m$ separate equations. A verifier who directly checks all rows must inspect the constraints one by one. The QAP transformation compresses the list into a single polynomial divisibility statement.

Choose a constraint domain
\[
    H=\{r_1,r_2,\ldots,r_m\}\subset\F
\]
of $m$ distinct field elements, one point for each R1CS row. Its vanishing polynomial is
\[
    Z_H(X)=\prod_{i=1}^{m}(X-r_i).
\]
For every variable index $j\in\{0,\ldots,n-1\}$, interpolate three polynomials
\[
    A_j(X),\qquad B_j(X),\qquad C_j(X)
\]
from the $j$th columns of the R1CS matrices:
\[
    A_j(r_i)=A_{i,j},
    \qquad
    B_j(r_i)=B_{i,j},
    \qquad
    C_j(r_i)=C_{i,j}
    \qquad
    \text{for all }i.
\]
In words, the $j$th column of $A$ becomes the polynomial $A_j(X)$, the $j$th column of $B$ becomes $B_j(X)$, and the $j$th column of $C$ becomes $C_j(X)$.

Given an assignment $z=(z_0,\ldots,z_{n-1})$, combine these column polynomials using the assignment values as coefficients:
\[
    A_z(X)=\sum_{j=0}^{n-1}z_jA_j(X),
    \qquad
    B_z(X)=\sum_{j=0}^{n-1}z_jB_j(X),
    \qquad
    C_z(X)=\sum_{j=0}^{n-1}z_jC_j(X).
\]
At the constraint point $r_i$, interpolation gives
\[
    A_z(r_i)=\sum_j z_jA_j(r_i)=\sum_j z_jA_{i,j}=A_i\cdot z,
\]
with analogous equalities for $B_z(r_i)$ and $C_z(r_i)$. Therefore the $i$th R1CS constraint is equivalent to
\[
    A_z(r_i)B_z(r_i)-C_z(r_i)=0.
\]
Define the QAP numerator polynomial
\[
    P_z(X)=A_z(X)B_z(X)-C_z(X).
\]
Then the entire R1CS system is satisfied if and only if
\[
    P_z(r_i)=0\qquad\text{for every }r_i\in H.
\]
By the vanishing-polynomial criterion, this is equivalent to
\begin{tcolorbox}
\[
    Z_H(X)\mid P_z(X),
\]
or, equivalently, there exists a quotient polynomial $h_z(X)$ such that
\[
    A_z(X)B_z(X)-C_z(X)=h_z(X)Z_H(X).
\]
\end{tcolorbox}
This is the QAP relation. It is the exact algebraic statement that later pairing-based SNARKs prove succinctly.

\subsection{Where Groth16 begins}

At the QAP stage, computation correctness has become the claim that the prover knows an assignment $z$ and a quotient polynomial $h_z$ satisfying
\[
    A_z(X)B_z(X)-C_z(X)=h_z(X)Z_H(X).
\]
Groth16 begins by evaluating this relation at a hidden setup point $\tau$, encoding the resulting values in a structured reference string, and using pairings to check the multiplicative relation without revealing the witness. The proving and verification equations, public-input separation, toxic-waste structure, and zero-knowledge blinding terms will be developed in a later iteration.

\section{Synthesis and Outlook}

The conceptual chain across the preceding sections is now visible:
\begin{enumerate}[leftmargin=2em]
    \item Start with the goal of proving knowledge of a witness for a circuit.
    \item Encode the circuit and witness algebraically.
    \item Reduce global correctness to a handful of polynomial identities.
    \item Commit to the polynomials succinctly.
    \item Check those identities at random points and make the interaction non-interactive via Fiat-Shamir.
\end{enumerate}

At this point one can already understand the architecture of a large fraction of modern proof systems. Later topics can explore different instantiations of the same template: different arithmetizations, different commitment schemes, different recursion techniques, and different implementation tradeoffs. The R1CS/QAP route developed above is the arithmetization layer for Groth16. The next technical layer to be developed is the pairing-based proof protocol itself: the structured reference string, the prover's group elements, the verifier's pairing equation, and the zero-knowledge blinding terms~\cite{Groth2016}.

\todo{merge the two}
\input{groth16}